### 2022-11-07 lmcs/conclusion.tex
restructuring before reworking
polishing
Created TODO.md, started fixes

% !TeX root = priorities.tex
\documentclass[main.tex]{subfiles}

\begin{document}

@@

\subsubsection*{Deadlock freedom and progress}
Deadlock freedom and progress are well studied properties in the $\pi$-calculus.
For the `standard' typed $\pi$-calculus, an important line of work starts from Kobayashi's approach to deadlock freedom~\cite{kobayashi98}, where priorities are values from an abstract poset. Kobayashi~\cite{kobayashi06} simplifies the abstract poset to pairs of naturals, called \emph{obligations} and\emph{capabilities}. Padovani simplifies these further to a single natural, called a \emph{priority}~\cite{padovani14}, and adapts obligations/capabilities to session types~\cite{padovani13}.

@@

\subsubsection*{Acknowledgements}
The authors would like to thank Simon Fowler, April Gon\c{c}alves, and Philip Wadler for their comments on the manuscript.

\end{document}

>>>>

@@

\subsubsection*{Deadlock freedom and progress}
Deadlock freedom and progress are well studied properties in the $\pi$-calculus.
For the `standard' typed $\pi$-calculus, an important line of work starts from Kobayashi's approach to deadlock freedom~\cite{kobayashi98}, where priorities are values from an abstract poset. Kobayashi~\cite{kobayashi06} simplifies the abstract poset to pairs of naturals, called \emph{obligations} and \emph{capabilities}. Padovani simplifies these further to a single natural, called a \emph{priority}~\cite{padovani14}, and adapts obligations/capabilities to session types~\cite{padovani13}.

@@

% \subsubsection*{Acknowledgements}
% The authors would like to thank Simon Fowler, April Gon\c{c}alves, and Philip Wadler for their comments on the manuscript.

### 2022-11-07 lmcs/introduction.tex
Created TODO.md, started fixes

In the $\pi$-calculus literature, there have been several attempts at developing Curry-Howard correspondences between session-typed $\pi$-calculus and linear logic~\cite{girard87}: Caires and Pfenning's $\pi$DILL~\cite{cairespfenning10} corresponds to dual intuitionistic linear logic~\cite{barber96}, and Wadler's Classical Processes~\cite[CP]{wadler14} corresponds to classical linear logic~\cite[CLL]{girard87}. Both calculi guarantee deadlock freedom, which they achieve by restricting structure of processes and shared channels to \emph{trees}, by combing name restriction and parallel composition into a single construct, corresponding to the logical cut. This ensures that two processes can only communicate via exactly one series of channels, which rules out interleavings of sessions, and guarantees deadlock freedom.
There are many downsides to combining name restriction and parallel composition, such as lack of modularity, difficulty typing structural congruence and formulating label-transition semantics, which have led to various approaches to decoupling these constructs. Hypersequent CP~\cite{MP18,kokkemontesi19popl,kokkemontesi19tlla} and Linear Compositional Choreographies~\cite{CarboneMS18} decouple them, but maintain the correspondence to CLL and allow only tree-structured processes. Priority CP~\cite[PCP]{dardhagay18extended} weakens the correspondence to CLL in exchange for a more expressive language which allows cyclic-structured processes. PCP decouples CP's cut rule into two separate constructs: one for parallel composition via a mix rule, and one for name restriction via a cycle rule. To restore deadlock freedom, PCP uses \emph{priorities}~\cite{kobayashi06,padovani14}. Priorities encode the \emph{order of actions} and rule out bad cyclic interleavings. Dardha and Gay~\cite{dardhagay18extended} prove cycle-elimination for PCP, adapting the cut-elimination proof for classical linear logic, and deadlock freedom follows as a corollary.

>>>>

In the $\pi$-calculus literature, there have been several attempts at developing Curry-Howard correspondences between session-typed $\pi$-calculus and linear logic~\cite{girard87}: Caires and Pfenning's $\pi$DILL~\cite{cairespfenning10} corresponds to dual intuitionistic linear logic~\cite{barber96}, and Wadler's Classical Processes~\cite[CP]{wadler14} corresponds to classical linear logic~\cite[CLL]{girard87}. Both calculi guarantee deadlock freedom, which they achieve by restricting structure of processes and shared channels to \emph{trees}, by combining name restriction and parallel composition into a single construct, corresponding to the logical cut. This ensures that two processes can only communicate via exactly one series of channels, which rules out interleavings of sessions, and guarantees deadlock freedom.
There are many downsides to combining name restriction and parallel composition, such as lack of modularity, difficulty typing structural congruence and formulating label-transition semantics, which have led to various approaches to decoupling these constructs. Hypersequent CP~\cite{MP18,kokkemontesi19popl,kokkemontesi19tlla} and Linear Compositional Choreographies~\cite{CarboneMS18} decouple them, but maintain the correspondence to CLL and allow only tree-structured processes. Priority CP~\cite[PCP]{dardhagay18extended} weakens the correspondence to CLL in exchange for a more expressive language which allows cyclic-structured processes. PCP decouples CP's cut rule into two separate constructs: one for parallel composition via a mix rule, and one for name restriction via a cycle rule. To restore deadlock freedom, PCP uses \emph{priorities}~\cite{kobayashi06,padovani14}. Priorities encode the \emph{order of actions} and rule out bad cyclic interleavings. Dardha and Gay~\cite{dardhagay18extended} prove cycle-elimination for PCP, adapting the cut-elimination proof for classical linear logic, and deadlock freedom follows as a corollary.

### 2022-11-07 lmcs/priority-gv.tex
Created TODO.md, started fixes

\begin{figure}
\begin{minipage}{0.5\linewidth}
\begin{example}[Cyclic Structure]%
  \label{ex:cycle}
  \[\tm{%
    \begin{array}{l}
      \letpair{x}{x'}{\new}{}
      \\
      \letpair{y}{y'}{\new}{}
      \\
      \spawn\left(%
      \begin{array}{l}
        {\letpair{\unit}{x'}{\recv\;{x'}}{}}
        \\
        {\letbind{y}{\send\;{\pair{\unit}{y}}}{}}
        \\
        {\andthen{\wait\;{x'}}{\close\;{y}}}
      \end{array}%
      \right);
      \\
      \underline{\letbind{x}{\send\;{\pair{\unit}{x}}}{}}
      \\
      \underline{\letpair{\unit}{y'}{\recv\;{y'}}{}}
      \\
      {\andthen{\close\;{x}}{\wait\;{y'}}}
    \end{array}
  }\]
\end{example}
\end{minipage}%
\begin{minipage}{0.5\linewidth}
\begin{example}[Deadlock]%
  \label{ex:deadlock}
  \[\tm{%
    \begin{array}{l}
      \letpair{x}{x'}{\new}{}
      \\
      \letpair{y}{y'}{\new}{}
      \\
      \spawn\left(%
      \begin{array}{l}
        {\letpair{\unit}{x'}{\recv\;{x'}}{}}
        \\
        {\letbind{y}{\send\;{\pair{\unit}{y}}}{}}
        \\
        {\andthen{\wait\;{x'}}{\close\;{y}}}
      \end{array}%
      \right);
      \\
      \underline{\letpair{\unit}{y'}{\recv\;{y'}}{}}
      \\
      \underline{\letbind{x}{\send\;{\pair{\unit}{x}}}{}}
      \\
      {\andthen{\close\;{x}}{\wait\;{y'}}}
    \end{array}
   }\]
\end{example}
\end{minipage}
\end{figure}

@@

\subsection{Operational Semantics}
\subsubsection*{Configurations}
Priority GV terms are evaluated as part of a configuration of processes. Configurations are defined by the following grammar:
\[
  \tm{\phi}
  \Coloneqq \tm{\main}
  \sep      \tm{\child}
  \hfill
  \tm{\conf{C}}, \tm{\conf{D}}, \tm{\conf{E}}
  \Coloneqq \tm{\phi\;M}
  \sep      \tm{\ppar{\conf{C}}{\conf{D}}}
  \sep      \tm{\res{x}{x'}{\conf{C}}}
  \hfill
\]

Configurations ($\tm{\conf{C}}$, $\tm{\conf{D}}$, $\tm{\conf{E}}$) consist of threads $\tm{\phi\;M}$, parallel compositions $\tm{\ppar{\conf{C}}{\conf{D}}}$, and name restrictions $\tm{\res{x}{x'}{\conf{C}}}$. To preserve the functional nature of PGV, where programs return a single value, we use flags ($\tm{\phi}$) to differentiate between the main thread, marked $\tm{\main}$, and child threads created by $\tm{\spawn}$, marked $\tm{\child}$. Only the main thread returns a value. We determine the flag of a configuration by combining the flags of all threads in that configuration:
\[
  \tm{\main}  + \tm{\child} = \tm{\main}
  \hfill
  \tm{\child} + \tm{\main}  = \tm{\main}
  \hfill
  \tm{\child} + \tm{\child} = \tm{\child}
  \hfill
  (\tm{\main}  + \tm{\main} \; \text{is undefined})
  \hfill
\]

@@

Structural congruence satisfies the following axioms: $\LabTirName{SC-LinkSwap}$ allows swapping channels in the link process. $\LabTirName{SC-ResLink}$ allows restriction to applied to link which is structurally equivalent to the terminated process, thus allowing elimination of unnecessary restrictions. $\LabTirName{SC-ResSwap}$ allows swapping channels and $\LabTirName{SC-ResComm}$ states that restriction is commutative. $\LabTirName{SC-ResExt}$ is the standard scope extrusion rule. Rules $\LabTirName{SC-ParNil}$, $\LabTirName{SC-ParComm}$ and $\LabTirName{SC-ParAssoc}$ state that parallel composition uses the terminated process as the neutral element; it is commutative and associative.

@@

Intuitively, evaluating $\tm{\link}$ causes a substitution: if we have a channel bound as $\tm{\res{x}{y}{}}$, then $\tm{\link\;\pair{w}{x}}$ replaces all occurrences of $\tm{y}$ by $\tm{w}$. However, in GV, $\tm{\link}$ is required to return a terminated channel, which means that the semantics for $\tm{link}$ must \emph{create} a fresh channel of type $\ty{\tyends}/\ty{\tyendr}$. The endpoint of type $\ty{\tyends}$ is returned by the $\tm{link}$ construct, and a $\tm{\wait}$ on the other endpoint guards the \emph{actual} substitution. In PGV, evaluating $\tm{\link}$ simply causes a substitution.
\item
\emph{Our structural congruence is type preserving}. Consequently, we can embed it directly into the reduction relation. In GV, this is not the case, and subject reduction relies on proving that if $\equiv\cred$ ends up in an ill-typed configuration, we can rewrite it to a well-typed configuration using $\equiv$.
\end{enumerate}

@@

To illustrate this, let's go back to the deadlocked program in \cref{ex:deadlock}. Crucially, it composes the terms below in parallel. While each of these terms itself is well-typed, they impose opposite conditions on the priorities, so their composition is ill-typed. (We omit the priorities on $\ty{\tyends}$ and $\ty{\tyendr}$.)
\begin{mathpar}
  \inferrule*{
    \tseq[\cs{o'}]
    {\tmty{y'}{\tyrecv[\cs{o'}]{\tyunit}{\tyendr}}}
    {\recv\;{y'}}
    {\typrod{\tyunit}{\tyendr}}
    \\
    \tseq[\cs{p}]
    {\tmty{x}{\tysend[\cs{o}]{\tyunit}{\tyends}},\tmty{y'}{\tyendr}}
    {\letbind{x}{\send\;{\pair{\unit}{x}}}{\dots}}
    {\tyunit}
    \and
    \cs{o'}<\cs{o}
  }{%
    \tseq[\cs{p}]
    {%
      \tmty{x}{\tysend[\cs{o}]{\tyunit}{\tyends}},
      \tmty{y'}{\tyrecv[\cs{o'}]{\tyunit}{\tyendr}}
    }
    {\letpair{\unit}{y'}{\recv\;{y'}}{\letbind{x}{\send\;{\pair{\unit}{x}}}{\dots}}}
    {\tyunit}
  }

@@

\subsection{Progress and Deadlock Freedom}
PGV satisfies progress, as PGV configurations either reduce or are in normal form. However, the normal forms may seem surprising at first, as evaluating a well-typed PGV term does not necessarily produce \emph{just} a value. If a term returns an endpoint, then its normal form contains a thread which is ready to communicate on the dual of that endpoint. This behaviour is not new to PGV.

@@

A~configuration either reduces, or it is equivalent to configuration in normal form. Crucial to the normal form is that each term $\tm{M_i}$ is blocked on the corresponding channel $\tm{x_i}$, and hence no two terms act on dual endpoints. Furthermore, no term $\tm{M_i}$ can perform a communication action by itself, since those are excluded by the definition of actions.
Finally, as a corollary, we get that well-typed terms which do not return endpoints return \emph{just} a value:

>>>>

\begin{figure}
\begin{minipage}{0.5\linewidth}
\begin{example}[Cyclic Structure]%
  \label{ex:cycle}
  \[\tm{%
    \begin{array}{l}
      \letpair{x}{x'}{\new\unit}{}
      \\
      \letpair{y}{y'}{\new\unit}{}
      \\
      \spawn\left(\lambda\unit.%
      \begin{array}{l}
        {\letpair{\unit}{x'}{\recv\;{x'}}{}}
        \\
        {\letbind{y}{\send\;{\pair{\unit}{y}}}{}}
        \\
        {\andthen{\wait\;{x'}}{\close\;{y}}}
      \end{array}%
      \right);
      \\
      \underline{\letbind{x}{\send\;{\pair{\unit}{x}}}{}}
      \\
      \underline{\letpair{\unit}{y'}{\recv\;{y'}}{}}
      \\
      {\andthen{\close\;{x}}{\wait\;{y'}}}
    \end{array}
  }\]
\end{example}
\end{minipage}%
\begin{minipage}{0.5\linewidth}
\begin{example}[Deadlock]%
  \label{ex:deadlock}
  \[\tm{%
    \begin{array}{l}
      \letpair{x}{x'}{\new\unit}{}
      \\
      \letpair{y}{y'}{\new\unit}{}
      \\
      \spawn\left(\lambda\unit.%
      \begin{array}{l}
        {\letpair{\unit}{x'}{\recv\;{x'}}{}}
        \\
        {\letbind{y}{\send\;{\pair{\unit}{y}}}{}}
        \\
        {\andthen{\wait\;{x'}}{\close\;{y}}}
      \end{array}%
      \right);
      \\
      \underline{\letpair{\unit}{y'}{\recv\;{y'}}{}}
      \\
      \underline{\letbind{x}{\send\;{\pair{\unit}{x}}}{}}
      \\
      {\andthen{\close\;{x}}{\wait\;{y'}}}
    \end{array}
   }\]
\end{example}
\end{minipage}
\end{figure}

@@

\subsection{Operational Semantics}
\subsubsection*{Configurations}
Priority GV terms are evaluated as part of a configuration of processes. Configurations are defined by the following grammar:
\[
  \tm{\phi}
  \Coloneqq \tm{\main}
  \sep      \tm{\child}
  \hfill\qquad\hfill
  \tm{\conf{C}}, \tm{\conf{D}}, \tm{\conf{E}}
  \Coloneqq \tm{\phi\;M}
  \sep      \tm{\ppar{\conf{C}}{\conf{D}}}
  \sep      \tm{\res{x}{x'}{\conf{C}}}
\]

Configurations ($\tm{\conf{C}}$, $\tm{\conf{D}}$, $\tm{\conf{E}}$) consist of threads $\tm{\phi\;M}$, parallel compositions $\tm{\ppar{\conf{C}}{\conf{D}}}$, and name restrictions $\tm{\res{x}{x'}{\conf{C}}}$. To preserve the functional nature of PGV, where programs return a single value, we use flags ($\tm{\phi}$) to differentiate between the main thread, marked $\tm{\main}$, and child threads created by $\tm{\spawn}$, marked $\tm{\child}$. Only the main thread returns a value. We determine the flag of a configuration by combining the flags of all threads in that configuration:
\[
  \tm{\main}  + \tm{\child} = \tm{\main}
  \hfill\qquad\hfill
  \tm{\child} + \tm{\main}  = \tm{\main}
  \hfill\qquad\hfill
  \tm{\child} + \tm{\child} = \tm{\child}
  \hfill\qquad\hfill
  (\tm{\main}  + \tm{\main} \; \text{is undefined})
  \hfill\qquad\hfill
\]

@@

Structural congruence satisfies the following axioms: $\LabTirName{SC-LinkSwap}$ allows swapping channels in the link process. $\LabTirName{SC-ResLink}$ allows restriction to be applied to link which is structurally equivalent to the terminated process, thus allowing elimination of unnecessary restrictions. $\LabTirName{SC-ResSwap}$ allows swapping channels and $\LabTirName{SC-ResComm}$ states that restriction is commutative. $\LabTirName{SC-ResExt}$ is the standard scope extrusion rule. Rules $\LabTirName{SC-ParNil}$, $\LabTirName{SC-ParComm}$ and $\LabTirName{SC-ParAssoc}$ state that parallel composition uses the terminated process as the neutral element; it is commutative and associative.

@@

Intuitively, evaluating $\tm{\link}$ causes a substitution: if we have a channel bound as $\tm{\res{x}{y}{}}$, then $\tm{\link\;\pair{w}{x}}$ replaces all occurrences of $\tm{y}$ by $\tm{w}$. However, in GV, $\tm{\link}$ is required to return a terminated channel, which means that the semantics for $\tm{link}$ must \emph{create} a fresh channel of type $\ty{\tyends}/\ty{\tyendr}$. The endpoint of type $\ty{\tyends}$ is returned by the $\tm{link}$ construct, and a $\tm{\wait}$ on the other endpoint guards the \emph{actual} substitution. In PGV, evaluating $\tm{\link}$ simply causes a substitution.
\item
\emph{Our structural congruence is type preserving}. Consequently, we can embed it directly into the reduction relation. In GV, this is not the case, and subject reduction relies on proving that if $\equiv\cred$ ends up in an ill-typed configuration, we can rewrite it to a well typed configuration using $\equiv$.
\end{enumerate}

@@

To illustrate this, let's go back to the deadlocked program in \cref{ex:deadlock}. Crucially, it composes the terms below in parallel. While each of these terms itself is well typed, they impose opposite conditions on the priorities, so their composition is ill-typed. (We omit the priorities on $\ty{\tyends}$ and $\ty{\tyendr}$.)
\begin{mathpar}
  \inferrule*{
    \tseq[\cs{o'}]
    {\tmty{y'}{\tyrecv[\cs{o'}]{\tyunit}{\tyendr}}}
    {\recv\;{y'}}
    {\typrod{\tyunit}{\tyendr}}
    \\
    \tseq[\cs{p}]
    {\tmty{x}{\tysend[\cs{o}]{\tyunit}{\tyends}},\tmty{y'}{\tyendr}}
    {\letbind{x}{\send\;{\pair{\unit}{x}}}{\dots}}
    {\tyunit}
    \and
    \cs{o'}<\cs{o}
  }{%
    \tseq[\cs{p}]
    {%
      \tmty{x}{\tysend[\cs{o}]{\tyunit}{\tyends}},
      \tmty{y'}{\tyrecv[\cs{o'}]{\tyunit}{\tyendr}}
    }
    {\letpair{\unit}{y'}{\recv\;{y'}}{\letbind{x}{\send\;{\pair{\unit}{x}}}{\dots}}}
    {\tyunit}
  }

@@

\subsection{Progress and Deadlock Freedom}
PGV satisfies progress, as PGV configurations either reduce or are in normal form. However, the normal forms may seem surprising at first, as evaluating a well typed PGV term does not necessarily produce \emph{just} a value. If a term returns an endpoint, then its normal form contains a thread which is ready to communicate on the dual of that endpoint. This behaviour is not new to PGV.

@@

A~configuration either reduces, or it is equivalent to configuration in normal form. Crucial to the normal form is that each term $\tm{M_i}$ is blocked on the corresponding channel $\tm{x_i}$, and hence no two terms act on dual endpoints. Furthermore, no term $\tm{M_i}$ can perform a communication action by itself, since those are excluded by the definition of actions.
Finally, as a corollary, we get that well typed terms which do not return endpoints return \emph{just} a value:
